\chapter{Background}

\section{Bayesian inference}
\label{sec:bayesian-inference}

\subsection{Principles}

Bayesian inference provides a probabilistic framework for updating uncertainty about unknown quantities after observing data.

Let $(\Theta,\cF)$ denote the parameter space and let $D$ denote the observed data.
Assume that the statistical model
$\{P_\theta : \theta \in \Theta\}$
is dominated by a common measure $\mu$, with likelihood
$L(D \mid \theta)$.
Let $P_0$ denote a prior distribution on $(\Theta,\cF)$.
Whenever $P_0$ admits a density $\pi_0$ with respect to a reference measure $\nu$ on $\Theta$, we write

\begin{equation}
P_0(\dd\theta)
=
\pi_0(\theta)\,\nu(\dd\theta).
\end{equation}

The Bayesian model is then specified by the likelihood function
$L(D\mid\theta)$
and the prior density
$\pi_0(\theta)$.
The likelihood describes the information about $\theta$ contained in the data, while the prior represents information available before observing $D$.
Bayes' theorem combines these two sources of information through the posterior distribution

\begin{equation}
    \pi(\theta\mid D)
    =
    \frac{L(D\mid\theta)\pi_0(\theta)}
    {\int_{\Theta} L(D\mid t)\pi_0(t)\,\dd t}.
    \label{eq:bayes-theorem-background}
\end{equation}

The denominator in \eqref{eq:bayes-theorem-background} is the marginal likelihood, or evidence, and is given by

\begin{equation}
    m(D)
    =
    \int_{\Theta} L(D\mid t)\pi_0(t)\,\dd t.
    \label{eq:marginal-likelihood-background}
\end{equation}

When $m(D)<\infty$, the posterior distribution is well-defined and integrates to one.
In this sense, Bayes' theorem updates the prior distribution through the likelihood, with the marginal likelihood ensuring that the resulting posterior is a proper probability distribution on $\Theta$.

For any posterior-integrable function $h:\Theta\to\bR$, posterior summaries are often expressed through expectations of the form

\begin{equation}
    \bE_{\pi(\theta\mid D)}\qty[h(\theta)]
    =
    \int_{\Theta} h(\theta)\pi(\theta\mid D)\,\dd\theta.
    \label{eq:posterior-expectation-background}
\end{equation}

Examples include posterior means, posterior variances, marginal probabilities, credible intervals, and posterior predictive quantities.
The main computational challenge is that the posterior distribution is usually available only up to a normalising constant.

Indeed, many Bayesian algorithms only require the unnormalised posterior density

\begin{equation}
    \gamma(\theta)
    =
    L(D\mid\theta)\pi_0(\theta),
    \label{eq:unnormalised-posterior-background}
\end{equation}

because the marginal likelihood $m(D)$ does not depend on $\theta$.

However, the marginal likelihood is still important for model comparison, empirical Bayes procedures, and doubly intractable models.
In low-dimensional conjugate models, the integral in \eqref{eq:marginal-likelihood-background} may be available analytically.

In modern Bayesian models, this integral is often unavailable, and posterior expectations such as \eqref{eq:posterior-expectation-background} must be approximated numerically.
This motivates the use of simulation-based methods.

\subsection{Monte Carlo methods}

Monte Carlo methods approximate integrals by random sampling.

Let $\pi$ be a probability distribution on $(\Theta,\mathcal{F})$, and suppose that the quantity of interest is

\begin{equation}
    \mu
    =
    \bE_{\pi}\qty[h(\theta)]
    =
    \int_{\Theta} h(\theta)\pi(\theta)\,\dd\theta.
    \label{eq:mc-target-background}
\end{equation}

If $\theta_1,\dots,\theta_N$ are independent samples from $\pi$, then the Monte Carlo estimator of $\mu$ is

\begin{equation}
    \widehat{\mu}_N
    =
    \frac{1}{N}\sum_{i=1}^{N} h(\theta_i).
    \label{eq:mc-estimator-background}
\end{equation}

Under standard integrability conditions, the law of large numbers gives

\begin{equation}
    \widehat{\mu}_N
    \xrightarrow[N\to\infty]{a.s.}
    \mu.
    \label{eq:mc-lln-background}
\end{equation}

If $h(\theta)$ has finite variance under $\pi$, the central limit theorem gives

\begin{equation}
    \sqrt{N}\qty(\widehat{\mu}_N-\mu)
    \xrightarrow[N\to\infty]{d}
    \mathcal{N}\qty(0,\sigma_h^2),
    \label{eq:mc-clt-background}
\end{equation}

where

\begin{equation}
    \sigma_h^2
    =
    \operatorname{Var}_{\pi}\qty[h(\theta)].
    \label{eq:mc-variance-background}
\end{equation}

The important feature of the Monte Carlo approximation is that its convergence rate is typically $N^{-1/2}$ and does not directly depend on the dimension of $\Theta$.
This makes Monte Carlo methods attractive for Bayesian inference, where posterior distributions are often high-dimensional.

The difficulty is that independent sampling from $\pi(\theta\mid D)$ is rarely possible.
Markov chain Monte Carlo methods address this difficulty by constructing a Markov chain whose invariant distribution is the desired posterior distribution.

The samples are then dependent, but posterior expectations can still be estimated using ergodic averages.
If $(\theta_n)_{n\geq 0}$ is a Markov chain with invariant distribution $\pi$, then the MCMC estimator is

\begin{equation}
    \widehat{\mu}_N^{\operatorname{MCMC}}
    =
    \frac{1}{N}\sum_{n=1}^{N} h(\theta_n).
    \label{eq:mcmc-estimator-background}
\end{equation}

Under suitable irreducibility, recurrence, and integrability conditions, the ergodic theorem gives \cite{robert2004, meyn2009, polasek2012}

\begin{equation}
    \widehat{\mu}_N^{\operatorname{MCMC}}
    \xrightarrow[N\to\infty]{a.s.}
    \bE_{\pi}\qty[h(\theta)].
    \label{eq:mcmc-ergodic-theorem-background}
\end{equation}

The price paid for replacing independent samples with a Markov chain is autocorrelation.

When samples are correlated, the effective amount of information in $N$ iterations may be much smaller than $N$ independent draws.
This motivates the use of diagnostics such as effective sample size, Monte Carlo standard errors, trace plots, and comparisons across independent chains.
Another important practical issue is initialization.

A Markov chain usually starts from a distribution different from $\pi$, and therefore the early iterations may not be representative of the target distribution.
The usual practical response is to discard an initial portion of the chain, known as burn-in.

However, burn-in is an asymptotic correction rather than an exact one.
This distinction is central to the later discussion of exact and unbiased MCMC methods.

\section{Classical Markov chain Monte Carlo methods}

Markov chain Monte Carlo methods construct a transition kernel $P$ on $(\Theta,\mathcal{F})$ such that the posterior distribution $\pi$ is invariant.

This means that, for every measurable set $A\in\mathcal{F}$,

\begin{equation}
    \int_{\Theta} P(\theta,A)\pi(\theta)\,\dd\theta
    =
    \pi(A).
    \label{eq:invariance-background}
\end{equation}

Equivalently, if $\theta_n\sim\pi$ and $\theta_{n+1}\mid\theta_n\sim P(\theta_n,\cdot)$, then $\theta_{n+1}\sim\pi$.
Invariance alone does not guarantee convergence from arbitrary starting points.

For convergence, the Markov chain must also satisfy appropriate irreducibility and recurrence conditions.
A common sufficient condition used to establish invariance is detailed balance \cite{meyn2009}.

The transition kernel $P$ satisfies detailed balance with respect to $\pi$ if

\begin{equation}
    \pi(\theta)P(\theta,\dd\theta')
    =
    \pi(\theta')P(\theta',\dd\theta)
    \label{eq:detailed-balance-background}
\end{equation}

as measures on $\Theta\times\Theta$.

Detailed balance implies invariance and gives a convenient route for proving that a Markov chain targets the desired posterior distribution.
The two classical algorithms discussed in this section are Gibbs sampling and Metropolis--Hastings.

Gibbs sampling exploits conditional distributions that can be sampled exactly.
Metropolis--Hastings is more general and only requires evaluation of the target density up to a normalising constant.

Both algorithms are asymptotically exact in the sense that their ergodic averages converge to posterior expectations under suitable conditions.
They are not exact in the finite-time sense, because samples obtained after a finite number of iterations still depend on the initial state.

\subsection{Gibbs sampling}

Gibbs sampling is designed for posterior distributions whose full conditional distributions are available and easy to sample from.

Suppose that the parameter can be partitioned into $K$ blocks,

\begin{equation}
    \theta
    =
    \qty(\theta_1,\dots,\theta_K),
    \qquad
    \Theta
    =
    \Theta_1\times\cdots\times\Theta_K.
    \label{eq:gibbs-blocks-background}
\end{equation}

For each block $\theta_k$, define the complementary vector

\begin{equation}
    \theta_{-k}
    =
    \qty(\theta_1,\dots,\theta_{k-1},\theta_{k+1},\dots,\theta_K).
    \label{eq:gibbs-complement-background}
\end{equation}

The full conditional distribution of the $k$-th block is

\begin{equation}
    \pi(\theta_k\mid\theta_{-k},D).
    \label{eq:gibbs-full-conditional-background}
\end{equation}

A systematic-scan Gibbs sampler updates the blocks sequentially.
Starting from $\theta^{(n)}=\qty(\theta_1^{(n)},\dots,\theta_K^{(n)})$, one iteration produces $\theta^{(n+1)}$ by sampling

\begin{align}
    \theta_1^{(n+1)}
    &\sim
    \pi\qty(\theta_1\mid\theta_2^{(n)},\dots,\theta_K^{(n)},D),
    \label{eq:gibbs-step-1-background}
    \\
    \theta_2^{(n+1)}
    &\sim
    \pi\qty(\theta_2\mid\theta_1^{(n+1)},\theta_3^{(n)},\dots,\theta_K^{(n)},D),
    \label{eq:gibbs-step-2-background}
    \\
    &\vdots
    \nonumber
    \\
    \theta_K^{(n+1)}
    &\sim
    \pi\qty(\theta_K\mid\theta_1^{(n+1)},\dots,\theta_{K-1}^{(n+1)},D).
    \label{eq:gibbs-step-k-background}
\end{align}

Each update leaves the posterior distribution invariant because it samples exactly from a conditional distribution of the target.
The composition of these invariant updates also leaves the posterior distribution invariant.

The main advantage of Gibbs sampling is that every proposed block update is accepted automatically.
In this sense, the algorithm avoids the tuning of an acceptance probability.

On the other hand, the most relevant limitation is that the required full conditional distributions may not be available in closed form.
Even when they are available, strong posterior dependence between blocks can lead to slow mixing.

This is especially relevant in hierarchical models, latent variable models, and models with highly correlated parameters.
In such cases, blocking, reparameterization, or hybrid Gibbs-within-Metropolis strategies may be necessary.

\subsection{Metropolis--Hastings}
\label{subsec:metropolis}

The Metropolis--Hastings algorithm constructs a Markov chain using a proposal distribution and an accept-reject correction.
Let $\pi$ denote the target posterior density, known up to a normalising constant.

Let $q(\theta'\mid\theta)$ be a proposal density for moving from the current state $\theta$ to a proposed state $\theta'$.
Given the current state $\theta$, the algorithm first samples

\begin{equation}
    \theta'
    \sim
    q(\cdot\mid\theta).
    \label{eq:mh-proposal-background}
\end{equation}

The proposal is then accepted with probability

\begin{equation}
    \alpha(\theta,\theta')
    =
    \min\qty{
    1,
    \frac{\pi(\theta')q(\theta\mid\theta')}
    {\pi(\theta)q(\theta'\mid\theta)}
    }.
    \label{eq:mh-acceptance-background}
\end{equation}

If the proposal is accepted, the next state is $\theta'$.
If the proposal is rejected, the chain remains at $\theta$.
Thus,

\begin{equation}
    \theta_{n+1}
    =
    \begin{cases}
    \theta',
    &
    \text{with probability } \alpha(\theta_n,\theta'),
    \\
    \theta_n,
    &
    \text{otherwise}.
    \end{cases}
    \label{eq:mh-transition-background}
\end{equation}

The acceptance probability in \eqref{eq:mh-acceptance-background} depends only on ratios of the target density.
Therefore, if $\pi(\theta)=\gamma(\theta)/Z$ for an unknown normalising constant $Z$, then $Z$ cancels from the ratio.

This cancellation is one of the main reasons why Metropolis--Hastings is useful in Bayesian computation.
The Metropolis--Hastings transition kernel satisfies detailed balance with respect to $\pi$.

Indeed, the accept-reject correction symmetrizes the probability flow between $\theta$ and $\theta'$.
Consequently, $\pi$ is invariant for the Markov chain.

A common special case is the random-walk Metropolis algorithm.
In this case,

\begin{equation}
    q(\theta'\mid\theta)
    =
    q(\theta-\theta'),
    \label{eq:random-walk-proposal-background}
\end{equation}

so that $q(\theta'\mid\theta)=q(\theta\mid\theta')$.
The acceptance probability then reduces to

\begin{equation}
    \alpha(\theta,\theta')
    =
    \min\qty{
    1,
    \frac{\pi(\theta')}{\pi(\theta)}
    }.
    \label{eq:random-walk-acceptance-background}
\end{equation}

The efficiency of Metropolis--Hastings depends strongly on the proposal distribution.
If the proposal variance is too small, the chain accepts many proposals but explores the state space slowly.

If the proposal variance is too large, the chain proposes distant states that are often rejected.
This trade-off is one of the central practical issues in MCMC.

Metropolis--Hastings is asymptotically exact under appropriate conditions, but it does not remove the finite-time dependence on the initial state.
For this reason, convergence diagnostics and burn-in choices remain important in practical applications.
In doubly intractable models, even the ratio in \eqref{eq:mh-acceptance-background} may contain unknown normalising constants that depend on $\theta$.

This is one of the settings where standard Metropolis--Hastings is no longer directly applicable and where the methods studied in this dissertation become necessary.

\section{Doubly intractable distributions}
\label{sec:dbl-int}

A posterior distribution is called \textit{doubly intractable} when its unnormalised
density cannot be computed pointwise \cite{murray2012, park2020}.
The difficulty is distinct from the ordinary intractability of the marginal likelihood
$m(D)$, which cancels from the Metropolis--Hastings ratio and therefore poses no
algorithmic obstacle.
In a doubly intractable model, the posterior density takes the form
\begin{equation}
    \label{eq:doubly-intractable-general}
    \pi(\theta \mid D)
    \propto
    \frac{\pi_0(\theta)\, h(D \mid \theta)}{Z(\theta)},
\end{equation}
where $Z(\theta) = \int h(D \mid \theta') \, \dd\theta'$ depends on the parameter $\theta$
and cannot be evaluated.
The ratio $Z(\theta) / Z(\theta')$ required by the Metropolis--Hastings acceptance step
is therefore itself intractable: both the numerator and the denominator involve the same
unknown normalising function evaluated at different parameter values.
This is the second source of intractability, in addition to the marginal likelihood.

Doubly intractable models arise naturally in several important classes.
Exponential random graph models have a likelihood of the form $\exp\{\theta^\top t(y)\} / Z(\theta)$,
where $Z(\theta)$ sums over all graphs of a given size and is computationally infeasible
for all but very small networks.
Gibbs random field models and Markov point processes share the same structure.

The normalised power prior, introduced in Section~\ref{sec:npp}, belongs to this class
for a different reason: here $Z(\theta)$ is not a partition function of the likelihood but
the normalising constant $c_0(\eta)$ of the \textit{prior} distribution, which depends on
the power parameter $\eta$.
This is an unusual member of the doubly intractable family, in that the intractable
constant arises through the Bayesian modelling choice rather than through the likelihood
structure itself.
The methods reviewed in Section~\ref{sec:mcmc-intractable} and developed in Chapter~\ref{sec:exact} are designed to handle precisely this setting.

\section{The Normalised Power Prior}
\label{sec:npp}

In many applied settings, a new study is conducted in a context where data from previous, related studies are already available.
Clinical trials are a canonical example: regulatory agencies and trial designers routinely face the question of how much weight to give historical control data when planning or analysing a new trial \cite{spiegelhalter2004, ibrahim2015}.
A natural Bayesian strategy is to use the historical data to construct an informative prior for the current analysis.
However, doing so naively, i.e., by treating the historical data as part of the current dataset, ignores the possibility that the two data sources are not fully exchangeable, for instance due to changes in patient populations, measurement protocols, or underlying disease dynamics.

The power prior framework of \citeonline{chen2000} addresses this by introducing a scalar parameter $\eta \in [0,1]$ that controls how much influence the historical data exert on the prior.
Rather than fully incorporating historical evidence, the prior is formed by raising the historical likelihood to the power $\eta$ before combining it with an initial prior.
When $\eta = 0$, the historical data are ignored entirely; when $\eta = 1$, they are weighted equally to a single observation from the current study.
Intermediate values allow a continuum of borrowing, and treating $\eta$ as unknown and inferring it from the data provides automatic, data-driven borrowing strength.

The normalised power prior \cite{duan2006, carvalho2021} is the version of the power prior that properly accounts for the normalising constant of the powered likelihood, ensuring that $\eta$ has a coherent joint prior interpretation alongside $\theta$.
As established in Section~\ref{sec:dbl-int}, this normalisation introduces a parameter-dependent constant $c_0(\eta)$ that renders the posterior doubly intractable.
This computational difficulty is the central challenge that this thesis addresses.

\subsection{Formulation}

Recalling the framework of Section~\ref{sec:bayesian-inference}, the posterior distribution of $\theta$ given data $D$ takes the form of \eqref{eq:bayes-theorem-background}, with $\pi_0(\theta)$ serving as the initial prior.
The normalised power prior framework replaces this initial prior by a data-informed version constructed from historical data $D_0 = \{d_{01}, \ldots, d_{0N_0}\}$.


\begin{definition}[Power prior]
    Let $D_0 = \{d_{01}, d_{02}, \dots, d_{0N_0}\}$ with $d_{0i} \in \mathcal{X}^{p}$ be the historical data, and let $L_0(D_0 \mid \theta)$ be a likelihood function that is finite for all $\theta \in \Theta \subseteq \bR^p$. Then, the power prior is defined as

    \begin{equation}
        \label{eq:power-prior}
        \pi_\eta(\theta \mid D_0) \propto L_0(D_0 \mid \theta)^{\eta} \pi_0(\theta),
    \end{equation}

    where $\pi_0(\theta)$ is the initial prior distribution of $\theta$ and $\eta$ is the (fixed) power parameter.
\end{definition}

To accommodate uncertainty in the choice of $\eta$, we treat it as a random variable and assign a prior distribution $\pi_A(\eta \mid \phi)$, indexed by hyperparameters $\phi$.

The unnormalised joint power prior over $(\theta, \eta)$ is then
\begin{equation}
    \pi^\star(\theta, \eta \mid D_0, \phi)
    \propto
    L_0(D_0 \mid \theta)^{\eta}\, \pi_0(\theta)\, \pi_A(\eta \mid \phi).
\end{equation}

To normalise this joint prior, one requires the normalising constant
\begin{equation}
    \label{eq:normalising-constant}
    c_0(\eta) = \int_{\Theta} L_0(D_0 \mid t)^{\eta}\, \dd P_0(t),
\end{equation}
which is unavailable in closed form for general likelihoods.

Given new data $D$, the marginal posterior distribution of $\eta$ is
\begin{equation}
    \label{eq:eta-marginal-posterior}
    \pi(\eta \mid D, D_0, \phi)
    \propto
    \frac{\pi_A(\eta \mid \phi)}{c_0(\eta)}
    \int_{\Theta} L(D \mid t)\, L_0(D_0 \mid t)^{\eta}\, \dd P_0(t).
\end{equation}

Since neither $c_0(\eta)$ nor the integral in \eqref{eq:eta-marginal-posterior} is tractable in general, this is a \textit{doubly intractable} posterior in the sense of Section~\ref{sec:dbl-int}.
The computational challenge this poses, and the strategies developed to address it, are the central subject of this thesis.

\subsection{The derivative of the normalising constant}
\label{subsec:deriv-normalising}

Under regularity conditions permitting differentiation under the integral sign, the thermodynamic identity (proved in Appendix~\ref{proof:deriv-normalising-constant}) gives
\begin{equation}
    \label{eq:normalising-constant-log}
    \dv{\log c_0(\eta)}{\eta}
    =
    \bE_{\pi_\eta}\bigl[\log L_0(D_0 \mid \theta)\bigr],
\end{equation}
where $\bE_{\pi_\eta}[\cdot]$ denotes expectation under the power prior $\pi_\eta(\theta \mid D_0)$.

This identity plays a central role in the exact inference methods of Chapter~\ref{sec:exact}: it expresses the log-derivative of an intractable normalising constant as a tractable posterior expectation, which can be estimated by Monte Carlo simulation.

\subsection{Other historical data modelling approaches}

The normalised power prior is one among several principled Bayesian approaches to incorporating historical information.
The \textit{commensurate prior} of \citeonline{hobbs2011} models the relationship between historical and current parameters through a shared hierarchical structure, allowing the degree of borrowing to adapt to the compatibility between studies.

\textit{Meta-analytic-predictive} (MAP) priors \cite{neuenschwander2010} derive an informative prior for the current study by fitting a hierarchical model to a collection of historical studies and marginalising over the between-study heterogeneity.

\textit{Robust mixture priors} \cite{schmidli2014} address the risk of prior-data conflict by mixing an informative component derived from historical evidence with a vague component, ensuring that the prior retains a non-negligible weight in regions not supported by the historical data.

Each of these alternatives makes different modelling assumptions about the exchangeability of the historical and current parameters, and each leads to different computational challenges.
The normalised power prior is particularly appealing when a single scalar parameter $\eta \in [0,1]$ suffices to capture the degree of historical relevance, and when the historical and current data share the same likelihood structure.

\section{MCMC methods for intractable normalising constants}
\label{sec:mcmc-intractable}

Many Bayesian models involve posterior distributions whose density can only be evaluated up to a normalising constant.
This is not an obstacle for ordinary Metropolis--Hastings when the unknown normalising constant does not depend on the parameter being updated.
However, the situation is considerably harder when the normalising constant itself depends on the parameter, as is the case for the normalised power prior.

In this section, we review three families of methods that are relevant for this thesis.
First, we discuss the exchange algorithm, which removes the intractable ratio of normalising constants by introducing an auxiliary draw from a proposal-dependent distribution.
Second, we discuss Barker's algorithm and Bernoulli factory methods, which replace the explicit evaluation of an acceptance probability by an exact randomisation procedure.
Third, we discuss divide-and-conquer Bernoulli factories, which aim to reduce the computational cost of Bernoulli factory methods in models with many likelihood factors.

\subsection{The exchange algorithm}
\label{subsec:exchange-algorithm}

Consider a family of unnormalised densities $f_\eta(\theta)$ with normalising constant
\begin{equation}
c(\eta)
=
\int_\Theta f_\eta(\theta) \dd \theta.
\end{equation}
Suppose that the target distribution has density
\begin{equation}
\pi(\eta \mid y)
\propto
\frac{\gamma(\eta \mid y)}{c(\eta)}.
\end{equation}
A standard Metropolis--Hastings proposal, as discussed in Section \ref{subsec:metropolis}, from $\eta$ to $\eta'$ requires the ratio
\begin{equation}
\frac{\pi(\eta' \mid y)}{\pi(\eta \mid y)}
=
\frac{\gamma(\eta' \mid y)}{\gamma(\eta \mid y)}
\frac{c(\eta)}{c(\eta')}.
\end{equation}
The ratio $c(\eta)/c(\eta')$ is generally unavailable.
This prevents the direct application of ordinary Metropolis--Hastings.

The exchange algorithm of \citeonline{murray2012} introduces an auxiliary variable $x^\star$ sampled exactly from the normalised model at the proposed value $\eta'$.
That is,
\begin{equation}
x^\star
\sim
\frac{f_{\eta'}(x)}{c(\eta')}.
\end{equation}
The Metropolis--Hastings ratio is then constructed on an augmented space so that the unknown normalising constants cancel.
In its simplest form, the exchange ratio is
\begin{equation}
R_{\mathrm{ex}}(\eta,\eta';x^\star)
=
\frac{\gamma(\eta' \mid y)}{\gamma(\eta \mid y)}
\frac{q(\eta \mid \eta')}{q(\eta' \mid \eta)}
\frac{f_\eta(x^\star)}{f_{\eta'}(x^\star)}.
\end{equation}
The corresponding acceptance probability is
\begin{equation}
\alpha_{\mathrm{ex}}(\eta,\eta')
=
\min
\left\{
1,
R_{\mathrm{ex}}(\eta,\eta';x^\star)
\right\}.
\end{equation}
The key point is that $R_{\mathrm{ex}}$ contains only unnormalised densities and proposal terms.
Therefore, the intractable ratio $c(\eta)/c(\eta')$ is not evaluated directly.

The exchange algorithm is particularly well-suited to settings where exact simulation from the normalised model is available.

\subsection{Barker's algorithm}

Barker's algorithm \cite{barker1965} is an alternative accept--reject rule for Markov chain Monte Carlo.
For a target distribution $\pi(\theta)$ and proposal density $q(\theta' \mid \theta)$, define the usual Metropolis--Hastings odds ratio as
\begin{equation}
R(\theta,\theta')
=
\frac{\pi(\theta')q(\theta \mid \theta')}
{\pi(\theta)q(\theta' \mid \theta)}.
\end{equation}
The Barker acceptance probability is
\begin{equation}
\alpha_B(\theta,\theta')
=
\frac{R(\theta,\theta')}
{1+R(\theta,\theta')}.
\end{equation}
Equivalently,
\begin{equation}
\alpha_B(\theta,\theta')
=
\frac{\pi(\theta')q(\theta \mid \theta')}
{\pi(\theta)q(\theta' \mid \theta)+\pi(\theta')q(\theta \mid \theta')}.
\end{equation}
This acceptance probability satisfies detailed balance because
\begin{equation}
\frac{\alpha_B(\theta,\theta')}
{\alpha_B(\theta',\theta)}
=
R(\theta,\theta').
\end{equation}
Therefore, Barker's algorithm leaves the target distribution invariant.

Barker's algorithm is usually less efficient than Metropolis--Hastings in the Peskun ordering 
\cite{goncalves2017}.
However, its acceptance probability has a useful algebraic form.
In particular, it can be implemented by Bernoulli factory methods when the odds ratio can be represented through probabilities that can be simulated but not evaluated.
This makes Barker's algorithm useful for some intractable likelihood problems \cite{goncalves2017,vats2021,}.

The important distinction is that Barker's algorithm is not required for the exchange algorithm.
The exchange algorithm can be combined with the ordinary Metropolis--Hastings acceptance probability.
Barker's rule becomes useful when the goal is to implement an accept--reject step through Bernoulli factories rather than through direct numerical evaluation of the acceptance probability.

\subsubsection{The two-coin algorithm}

The two-coin algorithm is a basic Bernoulli factory construction used to implement Barker-type accept--reject probabilities \cite{goncalves2017}.
Suppose that the odds ratio can be factorized as
\begin{equation}
R(\theta,\theta')
=
\frac{c(\theta',\theta)p(\theta',\theta)}
{c(\theta,\theta')p(\theta,\theta')},
\end{equation}
where $c(\cdot,\cdot)$ is tractable and $p(\cdot,\cdot)$ is a probability that can be simulated.
Then Barker's acceptance probability can be written as
\begin{equation}
\alpha_B(\theta,\theta')
=
\frac{c(\theta',\theta)p(\theta',\theta)}
{c(\theta',\theta)p(\theta',\theta)+c(\theta,\theta')p(\theta,\theta')}.
\end{equation}
The two-coin algorithm generates a Bernoulli random variable with this probability without evaluating $p(\theta',\theta)$ or $p(\theta,\theta')$ numerically.

\begin{algorithm}[H]
    Flip a coin with probability proportional to $c(\theta',\theta)$ against $c(\theta,\theta')$\;
    \uIf{the first-stage coin selects the proposed direction}{
        Flip a coin with probability $p(\theta',\theta)$\;
        \If{the second-stage coin is heads}{
            accept the proposal\;
        }
    }
    \uElse{
        Flip a coin with probability $p(\theta,\theta')$\;
        \If{the second-stage coin is heads}{
            reject the proposal\;
        }
    }
    \If{no decision has been made}{
        return to the first step\;
    }
    \caption{Two-coin algorithm}
    \label{alg:twocoin}
\end{algorithm}

The algorithm loops until one of the simulated coins produces a decisive event.
Conditioning on the first decisive event gives the Barker probability.
Thus, the resulting accept--reject decision is exact even when the probabilities $p(\theta',\theta)$ and $p(\theta,\theta')$ are only available through simulation.

The practical difficulty is that the expected number of loops can be large.
This happens when the decisive events are rare.
In models with many likelihood factors, this cost can grow rapidly with the data size.

\subsection{Divide-and-conquer Bernoulli factory}

For models with independent observations, the likelihood often factorizes as
\begin{equation}
L(D \mid \theta)
=
\prod_{i=1}^n L_i(\theta).
\end{equation}
A Bernoulli factory implementation may then require simulating a coin whose probability is a product of many factor probabilities.
If the required probability is
\begin{equation}
p(\theta)
=
\prod_{i=1}^n p_i(\theta),
\end{equation}
then $p(\theta)$ can become very small as $n$ increases.
A direct two-coin implementation may therefore have an expected number of loops that grows exponentially in $n$.
This is the main computational bottleneck addressed by divide-and-conquer Bernoulli factories.

The divide-and-conquer Bernoulli factory of \citeonline{stumpf2025} reduces this cost by recursively merging coins associated with smaller groups of factors.
Instead of constructing one coin for the full product at once, the method builds a tree.
The leaves correspond to individual factors or small batches of factors.
The internal nodes merge the coins produced by their children.
The root returns the final Bernoulli variable required by the Barker accept--reject step.

\subsubsection{The coin merge procedure}

The basic merge identity can be described in terms of odds.
Suppose that two independent coins have success probabilities
\begin{equation}
r_1
=
\frac{h_1}{h_1+g_1}
\end{equation}
and
\begin{equation}
r_2
=
\frac{h_2}{h_2+g_2}.
\end{equation}
The goal is to construct a coin with success probability
\begin{equation}
r_0
=
\frac{h_1h_2}{h_1h_2+g_1g_2}.
\end{equation}
The merge procedure repeatedly flips the two child coins until they agree.
If both child coins return one, the merged coin returns one.
If both child coins return zero, the merged coin returns zero.
If the two child coins disagree, the procedure is repeated.

\begin{algorithm}[H]
    Flip independent coins with probabilities $r_1$ and $r_2$\;
    \eIf{both coins return the same value}{
        return the common value\;
    }{
        return to the first step\;
    }
    \caption{Coin merge algorithm}
    \label{alg:coinmerge}
\end{algorithm}

The probability that the merged coin returns one is
\begin{align}
\bP(\text{return }1)
&=
\frac{r_1r_2}
{r_1r_2+(1-r_1)(1-r_2)}
\\
&=
\frac{h_1h_2}
{h_1h_2+g_1g_2}.
\end{align}
Thus, the merge operation multiplies the odds represented by the two child coins.

\subsubsection{Hierarchical factorization}

Given $n$ factors, the DCBF organizes them in a binary tree.
At the leaves, each node represents either one likelihood factor or a small batch of likelihood factors.
At each internal node, the two child coins are merged using Algorithm~\ref{alg:coinmerge}.
The root coin represents the product of all odds contributions.

\begin{algorithm}[H]
    \SetKwProg{Fn}{function}{}{end}
    \Fn{FlipDCBF$(\mathcal F)$}{
        \If{$\mathcal F$ contains one leaf batch}{
            \KwRet FlipLeafCoin$(\mathcal F)$\;
        }
        Split $\mathcal F$ into two approximately balanced subsets $\mathcal F_L$ and $\mathcal F_R$\;
        $C_L \gets$ FlipDCBF$(\mathcal F_L)$\;
        $C_R \gets$ FlipDCBF$(\mathcal F_R)$\;
        \KwRet MergeCoins$(C_L,C_R)$\;
    }
    \caption{Divide-and-conquer Bernoulli factory}
    \label{alg:dcbf}
\end{algorithm}

The role of the leaf coin is model-specific.
In some applications, a leaf coin can be simulated exactly using a tractable bound.
In other applications, it requires an unbiased estimator or an auxiliary construction.
The exactness of the full DCBF depends on the exactness of every leaf coin and every merge operation.
If a leaf coin is only approximate, then the full DCBF is also approximate.

For the normalised power prior, the DCBF can be combined with an exchange construction.
The exchange step first removes the ratio $c_0(\eta)/c_0(\eta')$ by drawing $\theta^\star$ from the powered historical posterior at $\eta'$.
The remaining exchange odds can be factorized over historical observations or over batches.
The DCBF can then be used to simulate the Barker coin associated with this factorized exchange ratio.
This route is exact only when the auxiliary exchange draw is exact and the leaf coins are exact.

\subsubsection{Computational complexity}

The motivation for the DCBF is that a balanced tree can reduce the cost of simulating product-form Bernoulli probabilities.
The vanilla two-coin algorithm may have exponential cost in the number of factors.
By contrast, \cite{stumpf2025} show that divide-and-conquer Bernoulli factory MCMC can have polynomial complexity under suitable regularity conditions.
The precise polynomial degree depends on the construction of the leaf coins and on the availability of efficient unbiased estimators of the intractable likelihood ratio.

A simplified complexity statement is
\begin{equation}
\bE[\mathrm{Cost}]
=
O
\left(
4^\ell \rho(n,\ell)
\right),
\end{equation}
where $\ell$ is the depth of the tree and $\rho(n,\ell)$ is the expected cost of the leaf-level two-coin algorithm for a batch of size $n/2^\ell$.
When the posterior concentrates at rate $n^{-1/2}$ and the proposal is optimally scaled, suitable choices of $\ell$ can lead to polynomial rather than exponential complexity.
This is the main theoretical advantage of the divide-and-conquer construction.

The practical message is more cautious.
The DCBF is exact only if its component coins are exact.
It can be computationally expensive if the merge procedure requires many repeated attempts.
It also requires valid bounds or exact local constructions at the leaves.
For this reason, in this thesis the DCBF is treated as a theoretically motivated exact method under bounded-support assumptions, while the exchange Metropolis--Hastings update is used as the main exact benchmark when exact powered-posterior simulation is available.

\subsection{Pseudo-marginal Metropolis--Hastings}

Pseudo-marginal Metropolis--Hastings is a modification of the Metropolis--Hastings algorithm for targets whose unnormalised density cannot be evaluated exactly, but can be estimated without bias.
Let the target distribution be

\begin{equation}
    \pi(\theta)
    =
    \dfrac{\gamma(\theta)}{Z},
\end{equation}

where $Z$ is unknown and where $\gamma(\theta)$ is not available pointwise.
Assume that, for each $\theta$, we can simulate an auxiliary random variable $U \sim m_\theta(\dd u)$ and compute a non-negative estimator $\widehat{\gamma}(\theta, U)$ satisfying

\begin{equation}
    \bE_{U \sim m_\theta}
    \qty[
        \widehat{\gamma}(\theta, U)
    ]
    =
    \gamma(\theta).
\end{equation}

The pseudo-marginal method applies an ordinary Metropolis--Hastings algorithm on the extended space $(\theta, U)$.
The extended target is

\begin{equation}
    \widetilde{\pi}(\theta, u)
    =
    \dfrac{\widehat{\gamma}(\theta, u) m_\theta(u)}{Z}.
\end{equation}

This construction is valid because the marginal distribution of $\theta$ under $\widetilde{\pi}$ is the desired target distribution.
Indeed,

\begin{align}
    \int \widetilde{\pi}(\theta, u) \dd u
    &=
    \dfrac{1}{Z}
    \int
    \widehat{\gamma}(\theta, u) m_\theta(u) \dd u,
    \\
    &=
    \dfrac{\gamma(\theta)}{Z},
    \\
    &=
    \pi(\theta).
\end{align}

Given the current state $(\theta, U)$, the algorithm proposes $\theta' \sim q(\dd\theta' \mid \theta)$ and independently samples $U' \sim m_{\theta'}(\dd u')$.
The proposal is accepted with probability

\begin{equation}
    \alpha_{\mathrm{PM}}
    \qty{
        (\theta, U),
        (\theta', U')
    }
    =
    \min
    \qty{
        1,
        \dfrac{
            \widehat{\gamma}(\theta', U')
            q(\theta \mid \theta')
        }{
            \widehat{\gamma}(\theta, U)
            q(\theta' \mid \theta)
        }
    }.
\end{equation}

A crucial implementation detail is that the estimator associated with the current state must be stored and reused after rejection.
If the estimator at the current state were regenerated at every iteration, the resulting chain would generally no longer be the standard pseudo-marginal chain on the extended space.
Under the non-negativity and unbiasedness assumptions, the algorithm is exact in the sense that the marginal invariant distribution of $\theta$ is $\pi$.
This result is the central guarantee of the pseudo-marginal approach of \citeonline{andrieu2009}.

The method is especially useful when $\gamma(\theta)$ contains an intractable likelihood or an intractable normalising constant ratio for which an unbiased non-negative estimator is available.
In the context of normalised power priors, one may use pseudo-marginal ideas when the ratio involving $c_0(\eta)$ is replaced by a positive unbiased estimator of the corresponding ratio.

The main caveat is that unbiasedness alone is not sufficient for good computational performance.
If the estimator $\widehat{\gamma}(\theta, U)$ has high variance, the chain may become sticky because unusually large estimates at the current state are difficult to move away from.
This phenomenon can produce high autocorrelation, poor effective sample size, and long burn-in periods even though the invariant distribution is theoretically correct.\

Another caveat is that the estimator must be non-negative.
Signed unbiased estimators do not directly define a probability density on the extended space and therefore do not yield the standard pseudo-marginal guarantee.
Moreover, pseudo-marginal Metropolis--Hastings should not be confused with plugging an unbiased estimator into an arbitrary nonlinear acceptance probability.
For instance, replacing an exact ratio $R$ by an unbiased estimate $\widehat{R}$ inside Barker's probability $\widehat{R}/(1+\widehat{R})$ is generally biased, since

\begin{equation*}
    \bE\qty[\frac{\widehat{R}}{1+\widehat{R}}]
    \neq
    \frac{R}{1+R}.
\end{equation*}

Thus, pseudo-marginal Metropolis--Hastings provides an exact route when a non-negative unbiased estimator of the unnormalised target is available, but its practical usefulness depends strongly on the stability and variance of that estimator.

\section{Function emulation methods}
\label{sec:function-emulation}

A common strategy for handling an intractable function is to replace it by a statistical surrogate constructed from a finite number of evaluations at design points.
In the context of the normalised power prior, the target function is $g(\eta) = \log c_0(\eta)$, whose structural properties are established in Sections~\ref{subsec:deriv-normalising} and Appendix~\ref{proof:log-c0-convex}.
These properties motivate the use of emulators that can incorporate prior structural information rather than treating $g$ as an arbitrary smooth function.
Two families of emulators are particularly relevant here: Gaussian processes and shape-constrained additive models.
Both are used in the experiments of Chapter~\ref{chap:approximate}, where they are assessed against the exact methods of Chapter~\ref{sec:exact}.

\subsection{Gaussian processes}
\label{subsec:gp-emulation}

A Gaussian process (GP) is a distribution over functions such that any finite collection of function values follows a multivariate normal distribution \cite{rasmussen2006}.
Formally, one writes $g \sim \mathcal{GP}(m, k)$, where $m: [0,1] \to \bR$ is the mean function and $k: [0,1]^2 \to \bR$ is the covariance kernel.
Given evaluations $\{(\eta_i, g(\eta_i))\}_{i=1}^n$, the GP posterior is again a GP whose mean and covariance can be computed in closed form.
This posterior mean serves as the surrogate $\widehat{g}(\eta)$, and the posterior variance quantifies the emulation uncertainty at unobserved values of $\eta$.

A key feature of GP emulation in this setting is that, since differentiation is a linear operation, the derivative process $g'$ is also a Gaussian process. We discuss this in more detail in Section~\ref{subsec:shape-constraints}.

The main practical limitation of GP emulation is its computational cost: fitting a GP to $n$ design points requires $O(n^3)$ operations for the matrix factorisation, which becomes prohibitive as $n$ grows.
This constraint is not severe for the one-dimensional function $g: [0,1] \to \bR$ studied here, but would require sparse or inducing-point approximations in higher-dimensional settings.



\subsection{Generalised additive models}
\label{subsec:gam-emulation}

Generalised additive models (GAMs) \cite{hastie1986, wood2017} provide a flexible semiparametric framework for regression in which the response is modelled as a sum of smooth functions of the predictors.
For a scalar covariate $x$, a GAM takes the form
\begin{equation}
    f(x) = \beta_0 + \sum_{j=1}^{p} s_j(x),
\end{equation}
where each $s_j$ is an unknown smooth function represented through a spline basis expansion.
The basis functions are chosen to span a rich function space while remaining computationally tractable; common choices include thin-plate splines, cubic regression splines, and B-splines \cite{wood2017}.

Model fitting proceeds by penalised likelihood maximisation, where a roughness penalty is added to the log-likelihood to control the complexity of each smooth component and prevent overfitting.
The penalty is typically of the form $\lambda_j \int [s_j''(x)]^2 \dd x$, and the smoothing parameter $\lambda_j$ is selected by cross-validation or by restricted maximum likelihood (REML).
This yields a smooth, data-adaptive approximation to $f$ that does not require specifying a parametric form.

The main limitation of standard GAMs for function emulation is that they impose no qualitative constraints on the fitted function.
In settings where the target function is known to satisfy structural properties such as monotonicity or convexity, unconstrained smoothers may produce artefacts, such as oscillations, sign violations, or locally inconsistent curvature, particularly in regions with sparse design points.
This motivates the shape-constrained extensions discussed in Section~\ref{subsec:scam-emulation}.